---
title: "lab_06_introduction_to_inference"
subtitle: "Analysis of Environmental Data Lab 634"
author: "Emma Zimmerman Greenlee"
date: 2021-10-13
output: 
  html_document:
      toc: true
      toc_float: true
---
# Prelab
# Formula for the standard error of the mean
```{r}
require(palmerpenguins)

#x = penguins$bill_depth_mm
#is.na(x)
#sum(is.na(x))

#length(x) - sum(is.na(x))

#sqrt(length(x) - sum(is.na(x)))
#sd(x, na.rm = TRUE) / sqrt(length(x) - sum(is.na(x)))

sse_mean = function(x)
  {
    sd(x, na.rm = TRUE) / sqrt(length(x) - sum(is.na(x)))
  }

sse_mean(penguins$bill_depth_mm)
```

# The penguin data
```{r}
boxplot(flipper_length_mm ~ species, data = penguins)
```

# 2-species data
```{r}
dat_pen = subset(penguins, species != "Gentoo")
boxplot(flipper_length_mm ~ species, data = dat_pen)

dat_pen = droplevels(subset(penguins, species != "Gentoo"))
{
  par(mfrow = c(1, 2))
  boxplot(flipper_length_mm ~ species, data = penguins)
  boxplot(flipper_length_mm ~ species, data = dat_pen)
}
```

# Resampling with replacement
```{r}
# for reproducibility
set.seed(123)

flipper_shuffled = sample(penguins$flipper_length_mm, replace = TRUE)
par(mfrow = c(1, 2))
boxplot(flipper_length_mm ~ species, data = penguins)
boxplot(flipper_shuffled ~ penguins$species, xlab = "species")
```

# Classic t-test: Adelie and Chinstrap penguins
```{r}
t.test(dat_pen$flipper_length_mm ~ dat_pen$species)
```

# Two-sample resampling
```{r}
# Reset the random number generator state for reproduceablility
set.seed(1)
flipper_shuffled = sample(dat_pen$flipper_length_mm)

boxplot(flipper_shuffled ~ dat_pen$species)
```

# Classical test on resampled data
```{r}
t_test_1 = t.test(flipper_shuffled ~ dat_pen$species)
t_test_1
```

# Difference of means
```{r}
t_test = t.test(dat_pen$flipper_length_mm ~ dat_pen$species)
t_test

t_test$estimate

diff_observed = round(diff(t_test$estimate), digits = 3)
print(diff_observed, digits = 3)
```

# Using aggregate()
```{r}
agg_means = aggregate(
  flipper_length_mm ~ species, 
  data = dat_pen, 
  FUN = mean, 
  na.rm = TRUE)
diff_observed = diff(agg_means[, 2])

agg_means
diff_observed
```

# Sample sizes
```{r}
table(dat_pen$species)

n_1 = 68
n_2 = 152

dat_1 = sample(dat_pen$flipper_length_mm, n_1, replace = TRUE)
dat_2 = sample(dat_pen$flipper_length_mm, n_2, replace = TRUE)

diff_simulated = 
  mean(dat_1, na.rm = TRUE) - mean(dat_2, na.rm = TRUE)

print(c(observed = diff_observed, simulated = diff_simulated))
```

# Simulation function
```{r}
x = dat_pen$flipper_length_mm
n_1 = 68
n_2 = 152

dat_1 = sample(x, n_1, replace = TRUE)
dat_2 = sample(x, n_2, replace = TRUE)

diff_simulated = 
  mean(dat_1, na.rm = TRUE) - mean(dat_2, na.rm = TRUE)

# Turn the resampling code into one function
two_group_resample = function(x, n_1, n_2)
  {
    dat_1 = sample(x, n_1, replace = TRUE)
    dat_2 = sample(x, n_2, replace = TRUE)
  
    mean(dat_1, na.rm = TRUE) - mean(dat_2, na.rm = TRUE)
  }
set.seed(54321)
two_group_resample(dat_pen$flipper_length_mm, 68, 152)
```

# Resampling experiment
```{r}
n = 2000
mean_differences = c()
for (i in 1:n)
{
  mean_differences = c(
    mean_differences,
    two_group_resample(dat_pen$flipper_length_mm, 68, 152)
  )
}
hist(mean_differences)

sum(abs(mean_differences) >= diff_observed)
```

# Retrieving named elements
```{r}
t_test = t.test(flipper_shuffled ~ dat_pen$species)
str(t_test)

t_test$estimate
```

# Lab Questions

# Question 1
```{r}
rm(list = ls())

sse_mean = function(x)
  {
    sd(x, na.rm = TRUE) / sqrt(length(x) - sum(is.na(x)))
  }

sse_mean(penguins$body_mass_g)
sse_mean(mtcars$mpg)
```

# Question 2
```{r}
set.seed(54321)
two_group_resample = function(x, n_1, n_2)
  {
    dat_1 = sample(x, n_1, replace = TRUE)
    dat_2 = sample(x, n_2, replace = TRUE)
  
    mean(dat_1, na.rm = TRUE) - mean(dat_2, na.rm = TRUE)
  }
```

# Question 3
My function performs Monte Carlo resampling, it simulates a null hypothesis.


# Question 4
```{r}
dat_pen = droplevels(subset(penguins, species != "Gentoo"))
{
  par(mfrow = c(1, 2))
  boxplot(flipper_length_mm ~ species, data = penguins)
  boxplot(flipper_length_mm ~ species, data = dat_pen)
}

two_group_resample = function(x, n_1, n_2)
  {
    dat_1 = sample(x, n_1, replace = TRUE)
    dat_2 = sample(x, n_2, replace = TRUE)
  
    mean(dat_1, na.rm = TRUE) - mean(dat_2, na.rm = TRUE)
  }

n = 2000
mean_differences = c()
for (i in 1:n)
  {
  mean_differences = c(
    mean_differences,
    two_group_resample(dat_pen$flipper_length_mm, 68, 152)
    )
  }
hist(mean_differences, xlab = "Mean differences", main = "Histogram of resampled mean")
```

# Question 5
0 of the resampled differences of means were greater than 5.8. The R code I used to check the number of differences is below.
```{r}
agg_means = aggregate(
  flipper_length_mm ~ species, 
  data = dat_pen, 
  FUN = mean, 
  na.rm = TRUE)
diff_observed = diff(agg_means[, 2])
sum(abs(mean_differences) >= diff_observed)
```

# Question 6
You'd need to run at *least* 10 million simulations to see a difference in mean flipper length greater than or equal to 5.8 mm.

# Question 7
```{r}
boxplot(dat_pen$body_mass_g ~ dat_pen$species, xlab ="Species", ylab = "Body mass")
```

# Question 8
The group means are Adelie: 3700.662 and Chinstrap: 3733.088. The difference between the means is 32.42598. The code I used is below.
```{r}
agg_means = aggregate(
  body_mass_g ~ species, 
  data = dat_pen, 
  FUN = mean, 
  na.rm = TRUE)

agg_means
diff_crit = diff(agg_means[, 2]); diff_crit
```

# Question 9
The p-value is 0.5879, which means that if I repeatedly sampled body mass for Adelie and Chinstrap penguins, I'd expect to observe a difference in mean body mass of 32.4 g or greater in ~59% of the experiments.
```{r}
t.test(dat_pen$body_mass_g ~ dat_pen$species)
```

# Question 10 and 11
601 differences in means were greater than diff_crit (which is 32.42598).
```{r}
two_group_resample = function(x, n_1, n_2)
  {
    dat_1 = sample(x, n_1, replace = TRUE)
    dat_2 = sample(x, n_2, replace = TRUE)
  
    mean(dat_1, na.rm = TRUE) - mean(dat_2, na.rm = TRUE)
  }

n = 1000
mean_differences = c()
for (i in 1:n)
{
  mean_differences = c(
    mean_differences,
    two_group_resample(dat_pen$body_mass_g, 68, 152)
  )
}
agg_means
diff_crit
sum(abs(mean_differences) >= diff_crit)

hist(mean_differences, xlab = "Mean differences", main = "Histogram of simulation mean differences")
```