---
title: "Lab 5: Uncertainty, Samples, and Populations"
subtitle: "Analysis of Environmental Data Lab 634"
author: "Emma Zimmerman Greenlee"
date: 2021-10-06
output: 
  html_document:
      toc: true
      toc_float: true
---
# Prelab Walkthrough

# Create a Ricker function
```{r}
ricker_fun = function(x, a, b) 
{
  return(a * x * exp(-b * x))
}

curve(
  ricker_fun(x, 1, 1), 
  from = 0, to = 5, add = FALSE, 
  main = "Ricker function: a = 1, b = 1",
  ylab = "f(x)", xlab = "x")
```

# Create an exponential function
```{r}
exp_fun = function(x, a, b) 
{
  return(a * exp(-b * x))
}

curve(
  exp_fun(x, 0.3, 1/15), add = FALSE, from = 0,
  to = 50,
  ann = FALSE, axes = TRUE, ylab = "f(x)");
  box()
```

# Simulated data on a line
```{r}
# Seed the RNG so we can reproduce our results
set.seed(1234567)

# Specify the x-range and number of points:
n_pts = 50
x_min = 2
x_max = 10

# Generate the x-values
x_sim = runif(n_pts, min = x_min, max = x_max)

# Choose an intercept and slope for our deterministic model and generate the ‘predicted’ y values
param_intercept = 2.3
param_slope = 0.67
y_pred = param_intercept + x_sim * param_slope
plot(x_sim, y_pred, main = "Simulated Data\nNo Errors", xlab = "", ylab = "")
```

# Normal errors 1
# Add some normally-distributed noise to generate our 'observed' y-values
```{r}
error_mean = 0
error_sd = 0.25

y_observed = 
  y_pred + 
  rnorm(
    n = n_pts, 
    mean = error_mean, 
    sd = error_sd)
plot(x_sim, y_observed, main = "Normally Distributed Errors\n Constant Variance", xlab = "", ylab = "")
```

# Normal errors 2
# To use a more sophisticated stochastic model, we can make the variability larger with increasing values of x
```{r}
error_mean = 0
error_sd = 0.1

y_observed_2 = 
  y_pred + 
  rnorm(
    n = n_pts, 
    mean = error_mean, 
    sd = error_sd * x_sim)

par(mfrow = c(1, 2))
plot(x_sim, y_observed, main = "Normally Distributed Errors\n Constant Variance", xlab = "", ylab = "")
plot(x_sim, y_observed_2, main = "Normally Distributed Errors\n Increasing Variance", xlab = "", ylab = "")
```

# Choosing a model
```{r}
par(mfrow = c(3, 1))
plot(x_sim, y_observed)
plot(x_sim, y_observed_2)

par(mfrow = c(3, 1))
hist(y_observed - y_pred, main = "sim data 1", xlab = "observed y=values")
hist(y_observed_2 - y_pred, main = "sim data 2", xlab = "observed y=values")
```

# Lab 5
```{r}
# Load the data into a dataframe
require(here)

dat_dispersal = data.frame(read.csv(here("data", "dispersal.csv")))
head(dat_dispersal)
```

# Question 1
```{r}
exp_fun = function(x, a, b) 
{
  return(a * exp(-b * x))
}
```

# Question 2
```{r}
{
  curve(
    exp_fun(x, 1.9, 0.1), add = FALSE, from = 0,
    to = 50,
    ann = FALSE, axes = TRUE, col = "black", lty = "solid",
    ylab = "f(x)"); box()
  curve(
    exp_fun(x, 1.9, 0.3), add = TRUE, col = "black", lty = "dotted")
  curve(
    exp_fun(x, 1.2, 0.2), add = TRUE, col = "red", lty = "solid")
  curve(
    exp_fun(x, 1.2, 0.4), add = TRUE,  col = "red", lty = "dotted")
}
```

# Question 3
When I change parameter a, it seems to change the height of where the curve starts. The red lines have lower a values and they start much lower on the y-axis than where the black lines start.

# Question 4
When I change parameter b, it seems to change the depth of the curve, or rput another way, how much the line curves. The dotted lines with lower b values have "deeper" curves, whereas the solid lines with higher b values have "shallower" curves.

# Question 5
```{r}
{
  curve(
    ricker_fun(x, 25, 0.1), 
    from = 0, to = 10, add = FALSE, col = "black", lty = "solid",
    ylab = "f(x)", xlab = "x")
  curve(
    ricker_fun(x, 20, 0.2), 
    add = TRUE, col = "black", lty = "dotted")
  curve(
    ricker_fun(x, 10, 0.2), 
    add = TRUE, col = "black", lty = "dotted")
  curve(
    ricker_fun(x, 75, 0.3), 
    add = TRUE, col = "red", lty = "solid")
  curve(
    ricker_fun(x, 50, 0.3), 
    add = TRUE, col = "red", lty = "dotted")
  curve(
    ricker_fun(x, 40, 0.3), 
    add = TRUE, col = "red", lty = "dotted")
}
```

# Question 6
Varying the a parameter seems to change the height of the curve. The lines with higher a values have higher peaks than the lines with lower a values.

# Question 7
Varying the b parameter seems to change how large and flat the curve is. The lines with lower b values have much larger and more arched curves, wheras the lines with higher b values have smaller and flatter curves. 


# Question 8
My intercept values are x = 1465 and y = 0.05. My slope value is -0.0003.

I chose the values of my intercept parameters by using locator(1) that was suggested in the instructions. I chose my slope value by trying out many different values and essentially guessing until I found a slope that I thought fit the model well.

# Question 9
```{r}
line_point_slope = function(x, x1, y1, slope)
  {
    get_y_intercept = 
      function(x1, y1, slope) 
        return(-(x1 * slope) + y1)
    
    linear = 
      function(x, yint, slope) 
        return(yint + x * slope)
    
    return(linear(x, get_y_intercept(x1, y1, slope), slope))
  }

guess_x = 1465
guess_y = 0.05
guess_slope = -0.0003
  
plot(x = dat_dispersal$dist.class, y = dat_dispersal$disp.rate.ftb)
curve(line_point_slope(x, guess_x, guess_y, guess_slope), add = T)

dat_dispersal$predicted_lin = line_point_slope(dat_dispersal$dist.class, guess_x, guess_y, guess_slope)
```

# Question 10
My values are a = 1 and b = 0.0033. I set my a value as 1 becuase I wanted the function to start at the beginning of the x scale, and I essentially guessed for the b value by putting it lots of numberrs until I came up with a good visual fit.

# Question 11
```{r}
plot(x = dat_dispersal$dist.class, y = dat_dispersal$disp.rate.ftb)
curve(
    exp_fun(x, 1, 0.0033), add = TRUE)

dat_dispersal$predicted_exp = exp_fun(dat_dispersal$dist.class, 1, 0.0033)
```



# Question 12
My values are a = 0.007 and b = 0.0037. I started with 1/400 for both values and then adjusted the values incrementally from there until I got a good visual fit.

# Question 13
```{r}
plot(x = dat_dispersal$dist.class, y = dat_dispersal$disp.rate.ftb)
curve(
    ricker_fun(x, 0.007, 0.0037), add = TRUE)

dat_dispersal$predicted_rick = ricker_fun(dat_dispersal$dist.class, 0.007, 0.0037)
```

# Question 14
```{r}
dat_dispersal$resids_linear = (dat_dispersal$disp.rate.ftb - dat_dispersal$predicted_lin)

dat_dispersal$resids_exp = (dat_dispersal$disp.rate.ftb - dat_dispersal$predicted_exp)

dat_dispersal$resids_ricker = (dat_dispersal$disp.rate.ftb - dat_dispersal$predicted_rick)

head(dat_dispersal)
```


# Question 15
```{r}
par(mfrow = c(3, 1))

hist(dat_dispersal$resids_linear)
hist(dat_dispersal$resids_exp)
hist(dat_dispersal$resids_ricker)
```
