---
title: "asgmt_frequentist_concepts"
subtitle: "Analysis of Environmental Data 602"
author: "Emma Zimmerman Greenlee"
date: 2021-10-20
output: 
  html_document:
      toc: true
      toc_float: true
---
# Question 1: 
### What is the probability of observing a count of exactly 3 successes in a binomial distribution with parameters n = 4 and p = 0.75?
```{r}
dbinom(3, size = 4, prob = 0.75, log = FALSE)
```
The probability is ~42%

# Question 2: 
### What is the probability of observing a count of 3 successes or fewer in a binomial distribution with parameters n = 4 and p = 0.75?
```{r}
pbinom(3, size = 4, prob = 0.75, log.p = FALSE)
```
The probability is ~68%

# Question 3: 
### What is the probability of observing more than 3 successes in a binomial distribution with parameters n = 5 and p = 0.75?
```{r}
1 - pbinom(4, size = 5, prob = 0.75, log.p = FALSE)
```
The probability is ~63%

# Question 4: 
### What is the probability of observing a value of less than 1.2 from a normally-distributed population with mean = 2 and standard deviation = 2?
```{r}
pnorm(1.2, mean = 2, sd = 2, log.p = FALSE)
```
The probability is ~34%

# Question 5: 
### What is the probability of observing a value of greater than 1.2 from a normally-distributed population with mean = 2 and standard deviation = 2?
```{r}
1 - pnorm(1.2, mean = 2, sd = 2, log.p = FALSE)
```
The probability is ~65%

# Question 6: 
### What is the probability of observing a value between 1.2 and 3.2 from a normally-distributed population with mean = 2 and standard deviation = 2?
```{r}
pnorm(3.2, mean = 2, sd = 2) - pnorm(1.2, mean = 2, sd = 2)
```
The probability is ~38%

# Question 7: 
### Describe how the shape of the histogram changes as you continue to press the sample button.
As I keep pressing the sample button, the histogram starts to resemble the skewed beta distribution more and more, height-wise.

# Question 8: 
### Describe how the shape of the histogram changes as you continue to press the sample button.
The histogram starts resembling the skewed beta distribution more, width-wise and a tail starts to appear, as well.

# Question 9: 
### Describe how the shape of the histogram changes as you continue to press the sample button.
Very little about the shape of the histogram changes as I press the sample button. The distribution of the data in the bars change slightly, but overall the shape stays the same.

# Question 10: 
### Why is there such a drastic change in the shape of the sampling distribution when you change the sample size from 1 to 2?
It's *only* taking the averages of the values, it's minimizing any/all outliers, and the data is, obviously, averaged instead of being raw.

# Question 11: 
### What are the two main factors that determine the width of the sampling distribution of the mean?
The standard deviation of the population and the sample size.

# Question 12: 
### How many 3-character words are possible?
25*25*25 = 25^3 = 15625 3-character words

# Question 13: 
### How many books would the Library contain if you added one additional position to the book size? Express your answer in terms of B.
B*25
