---
title: "Data exploration and deterministic functions"
subtitle: "Analysis of Environmental Data 602"
author: "Emma Zimmerman Greenlee"
date: 2021-09-16
output: 
  html_document:
      toc: true
      toc_float: true
---

# Question 1

```{r}

require("here")

dat_habitat = data.frame(read.csv(here("data", "hab.sta.csv")))

par(mfrow = c(3, 1))

hist(dat_habitat$elev, main = "Histogram of Elevation", xlab = "Elevation")

hist(dat_habitat$slope, main = "Histogram of Slope", xlab = "Slope")

hist(dat_habitat$aspect, main = "Histogram of Aspect", xlab = "Aspect")

```

# Question 2

The shape starts low and quickly gets taller and then drops off and tapers to an end as it gets higher. It is not evenly distributed, there are quite a bit more low elevation sampling sites

# Question 3

It's measured in percent slope.

# Question 4

The data goes up like a staircase and then plateaus for a bit and then quickly goes down. The sample sites are definitely not mostly flat, they are mostly steep slopes. There are not many shallow slopes.

# Question 5

Aspect is the orientation of a location/what direction an object is facing using the 360 degrees of a circle. 

# Question 6

The shape of the aspect histogram is mostly even all  the way across, with a sharp drop at the very end. Becuase the sampling sites are pretty evenly distributed they are all (more or less) equally facing each side.

# Question 7

```{r}
line_point_slope = function(x, x1, y1, slope)
{
  get_y_intercept = 
    function(x1, y1, slope) 
      return(-(x1 * slope) + y1)
  
  linear = 
    function(x, yint, slope) 
      return(yint + x * slope)
  
  return(linear(x, get_y_intercept(x1, y1, slope), slope))
}


par(mfrow = c(3, 1))

plot(x = dat_habitat$elev, 
     y = dat_habitat$ba.tot, 
     main = "Total BA to Elevation Scatterplot", 
     xlab = "Elevation", 
     ylab = "Total Basal Area")

curve(line_point_slope(x, x1 = mean(dat_habitat$elev), y1 = mean(dat_habitat$ba.tot), slope = 0.04), add = TRUE)

plot(x = dat_habitat$slope, 
     y = dat_habitat$ba.tot, 
     main = "Total BA to Slope Scatterplot", 
     xlab = "Slope", 
     ylab = "Total Basal Area")

curve(line_point_slope(x, x1 = mean(dat_habitat$slope), y1 = mean(dat_habitat$ba.tot), slope = 0.04), add = TRUE)

plot(x = dat_habitat$aspect, 
     y = dat_habitat$ba.tot, 
     main = "Total BA to Aspect Scatterplot", 
     xlab = "Aspect", 
     ylab = "Total Basal Area")

curve(line_point_slope(x, x1 = mean(dat_habitat$aspect), y1 = mean(dat_habitat$ba.tot), slope = 0.02), add = TRUE)
```


# Question 8

For the elevation variable, there is an association and it is linear. I think that although a linear model is a decent fit for this data, the hump in the middle of the data points makes it not a perfect fit.


For the slope variable, there again is a noticeable associations  and it is linear. A linear model is a good fit for this data because the data points are in a pretty straight line.


For the aspect variable, there is a noticeable association, and it is linear. I think that a linear model is a good fit for this data, as the data points are all in a relatively straight line with no curves up or down.


